Global code motion of congruent instructions

### 2017-04-18 paper.tex
use gcm / GCM macros

\def \GCC {GCC}
\def \LLVM {LLVM}
\def \CFG {CFG}
\def \SSA {SSA}
\def \MemorySSA {MemorySSA}
\def \PRE {PRE}
\def \GVN {GVN}
\def \SPEC {SPEC Cpu 2006}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a global code motion (GCM) compiler optimization which schedules
congruent instructions across the program.  Not only this saves code size, it
exposes redundancies in some cases, it exposes more instruction level
parallelism in the basic block to which instructions are moved, and it enables
other passes like loop invariant motion to remove more redundancies. The cost
model to drive the code motion is based on live range analysis on \SSA{}
representation such that the (virtual) register pressure does not increase, which
translates to 2\% fewer spills while compiling \SPEC{} benchmark suite.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on overall compile time on \SPEC{}. GCM enables more inlining and
exposes more loop invariant code motion opportunities in majority of the
benchmarks. We have also seen execution time improvements in a few of \SPEC{}
benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
\SPEC{} bencmarks suite when compiled for X86\_64. This is an optimistic
algorithm in the sense that it considers all congruent instructions in a
function as potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of code-motion, and how
code-motion increases instruction level parallelism thereby enabling more out of
order execution on modern architectures.
\end{abstract}

@@

This paper describes code-motion of congruent instructions \cite{briggs1997}, a
technique that uses the information computed for \PRE{} to detect identical
computations but has a transformation phase whose goal differs from \PRE{}: it
removes identical computations from different branches of execution.  These
identical computations in different branches of execution are not redundant
computations at run-time and the number of run-time computations is not
reduced. It is not a redundancy elimination pass, and thus it has different cost
function and heuristics than \PRE{} or CSE. It is more similar to global code
scheduling \cite{dragonbook,click1995global} in the sense that it will only move
computations. Code hoisting, for example, can reduce the critical path length of
execution in out-of-order machines. As more instructions are available at the
hoisting point, the hardware has more independent instructions to reorder. The
following example illustrates how hoisting can improve performance by exposing
more ILP.

@@

In fact, there are several advantages of GCM.
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single BB, should all the instructions in a conditional get
  hoisted or sinked;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP), which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored code-motion of congruent instructions in as much detail as we
have done. A part of our implementation (aggressive code hoisting) is already
merged in \LLVM{} trunk, however, a general implementation of code-motion of
congruent (non redundant) expressions is still missing from \GCC{} and \LLVM{}
trunk. The main contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

Dominance graphs, Global Value Numbers and Memory SSA are already available in
\LLVM{} but there is no facility to infer liveness of virtual registers. For
that we implemented DJ-Graph data structure that allows us to calculate
MergeSets which is used in the liveness analysis in our implementation of GCM.

@@

We compute merge sets of the control flow graph which remains the same through
the code-motion transformation. For code-motion we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

@@

The GCM pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for code-motion (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}), and
\item update data structures to continue iterative code motion.
\end{itemize}

@@

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a code-motion is profitable.
That takes into account the impact code-motion would have on various parameters
that might affect runtime performance, e.g., impact on live-range, gain in the
code size. Since this is mostly a code-size optimization pass, the goal is to
not regress in performance across popular benchmarks and at the same time to reduce
code size as much as possible. Following are the cost models which are
implemented:

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill), but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after code-motion is less than before it will
be moved. Essentially, as long as there is one killed operand, code hoisting
will either decrease or preserve the register pressure.  Similarly, code-sinking
will either decrease or preserve the register pressure as long as there is one
operand killed at most.  The following example explains how code motion of
identical computations can reduce the register pressure.

@@

Code-motion basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable, otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

@@

After no more candidates are hoistable, sinking is performed. Sinking is only
performed once becase currently there are very few sinkable candidates
(Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
GCM) has a code-sinking which already sinks several instructions.

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for code-motion, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ-graph and merge-sets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis expensive but only performed
on-demand for hoistable candidates so it does not affect the overall compile
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the code-motion, we have not seen signinficant
increase in the compilation times. We have also provided appropriate compiler
flags to expedite code-motion by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal (Section~\ref{sec:experimental-results}).

The analysis is followed by a simple code generation that adds the identified
instruction in the destination point and removes all instructions rendered
redundant as a result of code-motion.

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
We ran \SPEC{} with the patch. All the experiments were conducted on x86\_64
Ubuntu-Linux machine and at -Ofast optimization level. Each sub benchmark was
run three times and the best result was taken. We collected execution time,
code-size which is listed in Table~\ref{tab:code-size}, other compiler stats are
listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
Table~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time} to see the impact of code-motion.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
                &\% performance uplift & \% codesize decrease & \% codesize decrease\\
Benchmarks      & at -Ofast (high better)& at -Ofast (low better) & at -Os (low better)\\\hline
400.perlbench	& 1.00  & 1.00 & 1.00\\\hline
401.bzip2	& 1.00  & 1.00 & 1.00\\\hline
403.gcc	        & 1.00  & 1.01 & 1.00\\\hline
429.mcf	        & 1.03  & 1.00 & 1.00\\\hline
433.milc	& 1.00  & 1.00 & 1.00\\\hline
444.namd	& 1.00  & 1.01 & 1.00\\\hline
445.gobmk	& 0.99  & 1.01 & 1.00\\\hline
447.dealII	& 0.95  & 1.01 & 0.98\\\hline
450.soplex	& 0.99  & 0.99 & 1.00\\\hline
453.povray	& 0.97  & 1.00 & 1.00\\\hline
456.hmmer	& 0.99  & 1.00 & 1.00\\\hline
458.sjeng	& 1.02  & 1.00 & 1.00\\\hline
462.libquantum	& 1.00  & 1.00 & 1.00\\\hline
464.h264ref	& 1.00  & 1.00 & 1.00\\\hline
470.lbm	        & 1.00  & 1.00 & 1.00\\\hline
471.omnetpp	& 1.00  & 0.99 & 1.00\\\hline
473.astar	& 1.00  & 1.02 & 1.00\\\hline
482.sphinx3	& 0.99  & 1.00 & 1.00\\\hline
483.xalancbmk	& 1.01  & 0.99 & 1.00\\\hline
Geomean         & 0.997 & 1.00 & 1.00\\\hline
    \end{tabular}
  \end{center}
  \caption{Execution time and code size with and without code-motion on SPEC2006. Some performance uplift in mcf, and sjeng
  while code size improvement in dealII were observed.}
  \label{tab:code-size}
\end{table}

Overall there was only a minor change in performance regressed with a few ups
(429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
etc.). The code-motion pass was run twice in the pass pipeline (at -Ofast),
first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
local redundancies, GCM would have to analyze less redundant instructions, also
as inliner creates more redundancies, GCM would have more opportunities to move
congruent instructions out of branches.

Code size also remained almost the same as a result of code-motion as shown in
Table~\ref{tab:code-size}.  The code size listed here is the text size (from
linux size command) of the final executables for each benchmark.  Some bencmarks
like 473.astar increased in code size by almost 2\%. This is because once the
code-size of a function decreases (due to GCM), it becomes cheaper to inline. As
we can see in Table-\ref{tab:stats}, four more functions got inlined in
473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all
other benchmarks got more functions inlined.

The compile time metrics of GCM are listed in
Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
loads, stores and calls hoisted as well as removed. For each category, the
number of instructions removed is greater or equal to the number of instructions
hoisted because each hoisting is performed only when at least one identical
computation is found. Loads are hoisted the most followed by scalars, stores and
calls in decreasing order.  This was the common trend in all our
experiments. One reason why loads are hoisted the most is the early execution of
this pass (before mem2reg) in the \LLVM{} pass pipeline. Passes like mem2reg,
instcombine might actually remove those loads so this order may change should
this pass be scheduled later. The number of instructions sunk is very less
because \LLVM{} has a code sinking pass inside SimplifyCFG which runs before the
GCM and sinks some instructions. Another reason is that instructions for which
dependencies are not directly available in the successor
(Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking those
instructions would require a look ahead on the sinkability of the
dependencies. We plan to implement this in near future.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|r|r|r|}
      \hline
Number of spills  &base(-Ofast) &GCM(-Ofast) &\%change (low better) \\\hline
400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
401.bzip2	  & 718	        & 707	& 0.98 \\\hline
403.gcc	          & 5782	& 5722	& 0.99 \\\hline
429.mcf	          & 14	        & 17	& 1.21 \\\hline
433.milc	  & 635	        & 642	& 1.01 \\\hline
444.namd	  & 3223	& 3274	& 1.01 \\\hline
445.gobmk	  & 2166	& 2173	& 1.00 \\\hline
447.dealII	  & 11074	& 10234	& 0.92 \\\hline
450.soplex	  & 1122	& 1125	& 1.00 \\\hline
453.povray	  & 4701	& 4692	& 1.09 \\\hline
456.hmmer	  & 1197	& 1274	& 1.06 \\\hline
458.sjeng	  & 168	        & 176	& 1.05 \\\hline
462.libquantum	  & 118	        & 118	& 1.00 \\\hline
464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
470.lbm	          & 33	        & 33	& 1.00 \\\hline
471.omnetpp	  & 524	        & 527	& 1.00 \\\hline
473.astar	  & 180	        & 201	& 1.12 \\\hline
482.sphinx3	  & 674	        & 670	& 0.99 \\\hline
483.xalancbmk	  & 5102	& 4965	& 0.97 \\\hline
Grand Total	  & 43349	& 42497	& 0.98 \\\hline
    \end{tabular}
  \end{center}
  \caption{Number of spills before and after global-code-motion (GCM) on SPEC2006 at -Ofast}
  \label{tab:spills}
\end{table}

Spills are listed in Table~\ref{tab:spills}. Spills were reduced by an average
of 2\% on the \SPEC{} benchmark suite. This is a result of cost-model which
limits the code-motion to those candidates which do not increase the register
pressure (except for loads Section~\ref{subsec:cost-models}, overall spills
still decreases.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
  Benchmarks      &   Baseline	        & GCM           & \%Decrease (low better) \\\hline
  400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
  401.bzip2	  &   76,337,033	& 75,942,641	& 0.99 \\\hline
  403.gcc         &   1,079,625,231	& 1,073,158,220	& 0.99 \\\hline
  429.mcf         &   78,336,298	& 77,859,513	& 0.99 \\\hline
  433.milc	  &   272,525,306	& 269,684,876	& 0.99 \\\hline
  444.namd	  &   78,354,313	& 77,875,603	& 0.99 \\\hline
  445.gobmk	  &   261,229,078	& 258,459,798	& 0.99 \\\hline
  447.dealII	  &   643,457,797	& 638,628,331	& 0.99 \\\hline
  450.soplex	  &   245,442,755	& 242,811,500	& 0.99 \\\hline
  453.povray	  &   499,034,715	& 494,785,035	& 0.99 \\\hline
  456.hmmer	  &   214,425,324	& 211,984,389	& 0.99 \\\hline
  458.sjeng	  &   89,612,524	& 88,803,166	& 0.99 \\\hline
  462.libquantum  &   84,762,158	& 84,077,905	& 0.99 \\\hline
  464.h264ref	  &   151,049,588	& 149,284,624	& 0.99 \\\hline
  470.lbm         &   71,878,726	& 71,775,405	& 1.00 \\\hline
  471.omnetpp	  &   382,661,268	& 379,123,833	& 0.99 \\\hline
  473.astar	  &   78,351,117	& 77,872,820	& 0.99 \\\hline
  482.sphinx3	  &   158,434,290	& 156,586,406	& 0.99 \\\hline
  483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Change in total number of instructions executed during the
    compilation with and without global-code-motion (GCM) on \SPEC{} at -Ofast}
  \label{tab:compile-time}
\end{table}

The compilation time actually reduced for most of the benchmarks as a result of code motion.
This is because as code-motion removes (congruent) instructions, the number of instructions
to be processed by later compilation passes also reduces. The compilation time listed here
is the total instruction count as outputted from valgrind (--tool=cachegrind).

>>>>

\def \GCC {GCC}
\def \LLVM {LLVM}
\def \CFG {CFG}
\def \SSA {SSA}
\def \MemorySSA {MemorySSA}
\def \PRE {PRE}
\def \GVN {GVN}
\def \SPEC {SPEC Cpu 2006}
\def \gcm {global-code-motion}
\def \GCM {GCM}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a \gcm{} (\GCM{}) compiler optimization which schedules
congruent instructions across the program.  Not only this saves code size, it
exposes redundancies in some cases, it exposes more instruction level
parallelism in the basic block to which instructions are moved, and it enables
other passes like loop invariant motion to remove more redundancies. The cost
model to drive the code motion is based on live range analysis on \SSA{}
representation such that the (virtual) register pressure does not increase, which
translates to 2\% fewer spills while compiling \SPEC{} benchmark suite.

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on overall compile time on \SPEC{}. \GCM{} enables more inlining and
exposes more loop invariant code motion opportunities in majority of the
benchmarks. We have also seen execution time improvements in a few of \SPEC{}
benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
\SPEC{} bencmarks suite when compiled for X86\_64. This is an optimistic
algorithm in the sense that it considers all congruent instructions in a
function as potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of \gcm{}, and how
\gcm{} increases instruction level parallelism thereby enabling more out of
order execution on modern architectures.
\end{abstract}

@@

This paper describes \gcm{} of congruent instructions \cite{briggs1997}, a
technique that uses the information computed for \PRE{} to detect identical
computations but has a transformation phase whose goal differs from \PRE{}: it
removes identical computations from different branches of execution.  These
identical computations in different branches of execution are not redundant
computations at run-time and the number of run-time computations is not
reduced. It is not a redundancy elimination pass, and thus it has different cost
function and heuristics than \PRE{} or CSE. It is more similar to global code
scheduling \cite{dragonbook,click1995global} in the sense that it will only move
computations. Code hoisting, for example, can reduce the critical path length of
execution in out-of-order machines. As more instructions are available at the
hoisting point, the hardware has more independent instructions to reorder. The
following example illustrates how hoisting can improve performance by exposing
more ILP.

@@

In fact, there are several advantages of \GCM{}.
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on out-of-order processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single BB, should all the instructions in a conditional get
  hoisted or sinked;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP), which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored \gcm{} of congruent instructions in as much detail as we
have done. A part of our implementation (aggressive code hoisting) is already
merged in \LLVM{} trunk, however, a general implementation of \gcm{} of
congruent (non redundant) expressions is still missing from \GCC{} and \LLVM{}
trunk. The main contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructiosn out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

Dominance graphs, Global Value Numbers and Memory SSA are already available in
\LLVM{} but there is no facility to infer liveness of virtual registers. For
that we implemented DJ-Graph data structure that allows us to calculate
MergeSets which is used in the liveness analysis in our implementation of \GCM{}.

@@

We compute merge sets of the control flow graph which remains the same through
the \gcm{} transformation. For \gcm{} we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

@@

The \GCM{} pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for \gcm{} (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}), and
\item update data structures to continue iterative code motion.
\end{itemize}

@@

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \gcm{} is profitable.
That takes into account the impact \gcm{} would have on various parameters
that might affect runtime performance, e.g., impact on live-range, gain in the
code size. Since this is mostly a code-size optimization pass, the goal is to
not regress in performance across popular benchmarks and at the same time to reduce
code size as much as possible. Following are the cost models which are
implemented:

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill), but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after \gcm{} is less than before it will
be moved. Essentially, as long as there is one killed operand, code hoisting
will either decrease or preserve the register pressure.  Similarly, code-sinking
will either decrease or preserve the register pressure as long as there is one
operand killed at most.  The following example explains how code motion of
identical computations can reduce the register pressure.

@@

\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable, otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

@@

After no more candidates are hoistable, sinking is performed. Sinking is only
performed once becase currently there are very few sinkable candidates
(Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
\GCM{}) has a code-sinking which already sinks several instructions.

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ-graph and merge-sets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis is expensive and only performed
on-demand for hoistable candidates so it does not affect the overall compilation
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the \gcm{}, we have not seen signinficant
increase in the compilation times. We have also provided appropriate compiler
flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal (Section~\ref{sec:experimental-results}).

The analysis is followed by a simple code generation that adds the identified
instruction in the destination point and removes all instructions rendered
redundant as a result of \gcm{}.

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
We ran \SPEC{} with the patch. All the experiments were conducted on an
x86\_64-linux machine and at -Ofast optimization level. Each sub benchmark was
run three times and the best result was taken. We collected execution time,
code-size which is listed in Table~\ref{tab:code-size}, other compiler stats are
listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
Table~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time} to see the impact of \gcm{}.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
                &\% performance uplift & \% codesize decrease & \% codesize decrease\\
Benchmarks      & at -Ofast (high better)& at -Ofast (low better) & at -Os (low better)\\\hline
400.perlbench	& 1.00  & 1.00 & 1.00\\\hline
401.bzip2	& 1.00  & 1.00 & 1.00\\\hline
403.gcc	        & 1.00  & 1.01 & 1.00\\\hline
429.mcf	        & 1.03  & 1.00 & 1.00\\\hline
433.milc	& 1.00  & 1.00 & 1.00\\\hline
444.namd	& 1.00  & 1.01 & 1.00\\\hline
445.gobmk	& 0.99  & 1.01 & 1.00\\\hline
447.dealII	& 0.95  & 1.01 & 0.98\\\hline
450.soplex	& 0.99  & 0.99 & 1.00\\\hline
453.povray	& 0.97  & 1.00 & 1.00\\\hline
456.hmmer	& 0.99  & 1.00 & 1.00\\\hline
458.sjeng	& 1.02  & 1.00 & 1.00\\\hline
462.libquantum	& 1.00  & 1.00 & 1.00\\\hline
464.h264ref	& 1.00  & 1.00 & 1.00\\\hline
470.lbm	        & 1.00  & 1.00 & 1.00\\\hline
471.omnetpp	& 1.00  & 0.99 & 1.00\\\hline
473.astar	& 1.00  & 1.02 & 1.00\\\hline
482.sphinx3	& 0.99  & 1.00 & 1.00\\\hline
483.xalancbmk	& 1.01  & 0.99 & 1.00\\\hline
Geomean         & 0.997 & 1.00 & 1.00\\\hline
    \end{tabular}
  \end{center}
  \caption{Execution time and code size with and without \gcm{}
    (\GCM{}) on \SPEC{}: performance uplift at -Ofast in mcf and sjeng; code size
    improvement in dealII at -Os.}
  \label{tab:code-size}
\end{table}

Overall there was only a minor change in performance regressed with a few ups
(429.mcf, 458.sjeng etc.) and a few regresssions (447.dealII, 453.povray
etc.). The \gcm{} pass was run twice in the pass pipeline (at -Ofast),
first after EarlyCSE pass, and then after the inliner. Because EarlyCSE removes
local redundancies, \GCM{} would have to analyze less redundant instructions, also
as inliner creates more redundancies, \GCM{} would have more opportunities to move
congruent instructions out of branches.

Code size also remained almost the same as a result of \gcm{} as shown in
Table~\ref{tab:code-size}.  The code size listed here is the text size (from
linux size command) of the final executables for each benchmark.  Some bencmarks
like 473.astar increased in code size by almost 2\%. This is because once the
code-size of a function decreases (due to \GCM{}), it becomes cheaper to inline. As
we can see in Table-\ref{tab:stats}, four more functions got inlined in
473.astar. Except for 447.dealII, 450.soplex, 482.sphinx3 and 483.xalancbmk all
other benchmarks got more functions inlined.

The compile time metrics of \GCM{} are listed in
Table~\ref{tab:code-motion-metric}. The table lists the number of scalars,
loads, stores and calls hoisted as well as removed. For each category, the
number of instructions removed is greater or equal to the number of instructions
hoisted because each hoisting is performed only when at least one identical
computation is found. Loads are hoisted the most followed by scalars, stores and
calls in decreasing order.  This was the common trend in all our
experiments. One reason why loads are hoisted the most is the early execution of
this pass (before mem2reg) in the \LLVM{} pass pipeline. Passes like mem2reg,
instcombine might actually remove those loads so this order may change should
this pass be scheduled later. The number of instructions sunk is very less
because \LLVM{} has a code sinking pass inside SimplifyCFG which runs before the
\GCM{} and sinks some instructions. Another reason is that instructions for which
dependencies are not directly available in the successor
(Section~\ref{subsec:legality-sink}) is not sunk for now. Sinking those
instructions would require a look ahead on the sinkability of the
dependencies. We plan to implement this in near future.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|r|r|r|}
      \hline
Number of spills  &base(-Ofast) &\GCM{}(-Ofast) &\%change (low better) \\\hline
400.perlbench	  & 2539	& 2493	& 0.98 \\\hline
401.bzip2	  & 718	        & 707	& 0.98 \\\hline
403.gcc	          & 5782	& 5722	& 0.99 \\\hline
429.mcf	          & 14	        & 17	& 1.21 \\\hline
433.milc	  & 635	        & 642	& 1.01 \\\hline
444.namd	  & 3223	& 3274	& 1.01 \\\hline
445.gobmk	  & 2166	& 2173	& 1.00 \\\hline
447.dealII	  & 11074	& 10234	& 0.92 \\\hline
450.soplex	  & 1122	& 1125	& 1.00 \\\hline
453.povray	  & 4701	& 4692	& 1.09 \\\hline
456.hmmer	  & 1197	& 1274	& 1.06 \\\hline
458.sjeng	  & 168	        & 176	& 1.05 \\\hline
462.libquantum	  & 118	        & 118	& 1.00 \\\hline
464.h264ref	  & 3379	& 3454	& 1.02 \\\hline
470.lbm	          & 33	        & 33	& 1.00 \\\hline
471.omnetpp	  & 524	        & 527	& 1.00 \\\hline
473.astar	  & 180	        & 201	& 1.12 \\\hline
482.sphinx3	  & 674	        & 670	& 0.99 \\\hline
483.xalancbmk	  & 5102	& 4965	& 0.97 \\\hline
Grand Total	  & 43349	& 42497	& 0.98 \\\hline
    \end{tabular}
  \end{center}
  \caption{Number of spills before and after \gcm{}(\GCM{}) on SPEC2006 at -Ofast}
  \label{tab:spills}
\end{table}

Spills are listed in Table~\ref{tab:spills}. Spills were reduced by an average
of 2\% on the \SPEC{} benchmark suite. This is a result of cost-model which
limits the \gcm{} to those candidates which do not increase the register
pressure (except for loads Section~\ref{subsec:cost-models}, overall spills
still decreases.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{|l|c|c|c|}
      \hline
  Benchmarks      &   Baseline	        & \GCM{}           & \%Decrease (low better) \\\hline
  400.perlbench	  &   186,829,426	& 184,690,006	& 0.99 \\\hline
  401.bzip2	  &   76,337,033	& 75,942,641	& 0.99 \\\hline
  403.gcc         &   1,079,625,231	& 1,073,158,220	& 0.99 \\\hline
  429.mcf         &   78,336,298	& 77,859,513	& 0.99 \\\hline
  433.milc	  &   272,525,306	& 269,684,876	& 0.99 \\\hline
  444.namd	  &   78,354,313	& 77,875,603	& 0.99 \\\hline
  445.gobmk	  &   261,229,078	& 258,459,798	& 0.99 \\\hline
  447.dealII	  &   643,457,797	& 638,628,331	& 0.99 \\\hline
  450.soplex	  &   245,442,755	& 242,811,500	& 0.99 \\\hline
  453.povray	  &   499,034,715	& 494,785,035	& 0.99 \\\hline
  456.hmmer	  &   214,425,324	& 211,984,389	& 0.99 \\\hline
  458.sjeng	  &   89,612,524	& 88,803,166	& 0.99 \\\hline
  462.libquantum  &   84,762,158	& 84,077,905	& 0.99 \\\hline
  464.h264ref	  &   151,049,588	& 149,284,624	& 0.99 \\\hline
  470.lbm         &   71,878,726	& 71,775,405	& 1.00 \\\hline
  471.omnetpp	  &   382,661,268	& 379,123,833	& 0.99 \\\hline
  473.astar	  &   78,351,117	& 77,872,820	& 0.99 \\\hline
  482.sphinx3	  &   158,434,290	& 156,586,406	& 0.99 \\\hline
  483.xalancbmk	  &   22,485,708,759	& 22,455,060,569& 1.00 \\\hline
    \end{tabular}
  \end{center}
  \caption{Change in total number of instructions executed during the
    compilation with and without \gcm{} (\GCM{}) on \SPEC{} at -Ofast}
  \label{tab:compile-time}
\end{table}
